---
title: "Tanking"
format: html
---

# Roadmap

- I want to write an article about the tanking problem in the NBA, and show how it is affecting how each game is played after the trade deadline
- Some metrics/ideas I want to explore
  - Prevalence of blowouts (winning by 20+ points) before/after trade deadline
    - Average margin of victory calendar view
    - OR # of blowouts calendar view
  - Diverging winning percentages of playoff and non-playoff teams
  - Dramatic increase in minutes for "random players"
  - Star players playing less minutes, or less games on tanking teams.
  
## Calendar data wrangling

```{r}
library(tidyverse)
library(hoopR)
```

```{r}
seasons <- 2002:2026
dat <- load_nba_schedule(seasons = seasons)
dat_calendar <- dat%>%
  filter(season_type == 2)%>%
  mutate(mov = abs(home_score-away_score),
         blowout = if_else(mov >= 20, T, F),
         date = as_date(ymd_hm(date)))%>%
  group_by(date)%>%
  summarize(games_played = n(),
            avg_mov = mean(mov),
            blowouts = sum(blowout),
            season = min(season))


  
```

## Slope chart win% wrangling

### Helper function

```{r}
library(jsonlite)

extract_record <- function(x, record_type = "total") {
  x |>
    # Make it valid JSON
    gsub("'", '"', x = _) |>
    gsub("None", "null", x = _) |>
    gsub("\\}\n \\{", "},{", x = _) |>
    fromJSON() |>
    subset(type == record_type, select = summary) |>
    pull(summary)
}
```

### Main wrangling

```{r}
seasons <- 2002:2026
container <- vector("list", length = length(seasons))
for (i in seq_along(seasons)){
  container[[i]] <- espn_nba_standings(seasons[i])%>%
    mutate(year = seasons[i])
}
playoffs <- bind_rows(container)

playoffs <- playoffs%>%
  filter(year != 2020)%>%
  mutate(playoff_team = if_else(clincher == 4 | is.na(clincher), F, T))

playoffs_filtered <- playoffs%>%
  select(team, playoff_team, year)
```


```{r}
seasons <- 2021:2026
dat <- load_nba_schedule(seasons = seasons)

dat_cleaned <- dat%>%
  filter(status_type_name == "STATUS_FINAL",
         type_id == 1)%>%
   mutate(home_record = sapply(home_records, extract_record, record_type = "total"),
          away_record = sapply(away_records, extract_record, record_type = "total"),
          date = as_date(ymd_hm(start_date))) %>%
  separate(home_record, into = c("home_wins", "home_losses"), sep = "-", convert = TRUE)%>%
  separate(away_record, into = c("away_wins", "away_losses"), sep = "-", convert = TRUE)%>%
  select(date, home_display_name, away_display_name, home_wins, home_losses, away_wins, away_losses, season)%>%
  pivot_longer(
    cols = c(home_display_name, away_display_name),
    names_to  = "location",
    values_to = "team") %>%
  mutate(
    wins   = if_else(location == "home_display_name", home_wins,   away_wins),
    losses = if_else(location == "home_display_name", home_losses, away_losses)) %>%
  select(date, team, wins, losses, season)%>%
    arrange(team, date) %>%
  group_by(team, date, season) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  complete(team, date) %>%
  group_by(team) %>%
  fill(wins, losses, .direction = "down") %>%
  ungroup() %>%
  mutate(win_pct = wins / (wins + losses))

final_dat <- dat_cleaned%>%
  left_join(playoffs_filtered, by = join_by(team == team, season == year))
```


```{r}
seasons_2 <- c(2002:2019, 2021:2026)
dat_2 <- load_nba_schedule(seasons = seasons_2)
dat_2%>%
  filter(str_detect(home_display_name, "All-Stars"))
dat_3 <- dat_2%>%
  left_join(playoffs_filtered, by = join_by(home_display_name == team, season == year))%>%
  mutate(home_playoff_team = playoff_team)%>%
  select(-playoff_team)%>%
  left_join(playoffs_filtered, by = join_by(away_display_name == team, season == year))%>%
  mutate(away_playoff_team = playoff_team)%>%
  select(-playoff_team)%>%
  filter(status_type_name == "STATUS_FINAL",
         type_id == 1, 
         season_type == 2,
         home_id <=30)%>%
  mutate(date = as_date(ymd_hm(start_date)))%>%
  select(date, season, home_display_name, away_display_name, home_playoff_team, away_playoff_team, home_winner)%>%
  mutate(month = month(date),
         playoff_vs_non = if_else((home_playoff_team & !away_playoff_team) | (!home_playoff_team & away_playoff_team) , T, F),
         non_playoff_win = playoff_vs_non & ( (home_winner & !home_playoff_team ) | (!home_winner & home_playoff_team) ))%>%
  group_by(month, season)%>%
  summarize(total_playoff_non = sum(playoff_vs_non),
            wins_non = sum(non_playoff_win),
            win_pct = wins_non/total_playoff_non)%>%
  filter(total_playoff_non > 20)
  
dat_3%>%
  filter(season == 2026)%>%
  ggplot(aes(x = month, y = win_pct))+
  geom_col()+
  theme_minimal()
```


```{r}
library(slider)
final_dat_2 <- final_dat%>%
  mutate(games_played = wins + losses)%>%
  filter(games_played != 0,
         !is.na(games_played))%>%
  group_by(games_played, team, season)%>%
  summarize(win_pct = max(win_pct),
            playoff_team = max(playoff_team),
            wins = max(wins), losses = max(losses))%>%
  ungroup()%>%
  filter(!is.na(season))%>%
  group_by(team, season)%>%
  mutate(game_result = case_when(
      wins > lag(wins) ~ 1,                     # win
      TRUE ~ 0                                  # loss
    ),
    win_pct_10 = slide_dbl(
      game_result,
      mean,
      .before = 9,   # look back 10 games (9 before + current)
      .complete = TRUE,  # return NA until window is full
      na.rm = TRUE
    ))

final_dat_2%>%
  filter(!is.na(win_pct_10))%>%
  group_by(games_played, playoff_team, season)%>%
  summarize(avg_10 = mean(win_pct_10))%>%
  filter(season == 2025)%>%
  ggplot(aes(x = games_played, y = avg_10, group = playoff_team, color = playoff_team))+
  geom_line()+
  geom_vline(xintercept = 55, linetype = "dotted")+
  theme_minimal()

final_dat_3 <- final_dat_2%>%
  filter(!is.na(win_pct_10))%>%
  group_by(games_played, playoff_team, season)%>%
  summarize(avg_10 = mean(win_pct_10))%>%
  filter(season == 2026)
```


## Player minute wrangling

```{r}

seasons <- 2021:2026
trade_deadlines <- c(
  "2021-03-25",  # 2020-21 season
  "2022-02-10",  # 2021-22 season
  "2023-02-09",  # 2022-23 season
  "2024-02-08",  # 2023-24 season
  "2025-02-06",  # 2024-25 season
  "2026-02-05"   # 2025-26 season
)

deadline_df <- data.frame(season = seasons, deadline_date = trade_deadlines)

df <- load_nba_player_box(seasons = seasons)
playoff_joined <- df%>%
  filter(season_type == 2)%>%
  left_join(playoffs_filtered, join_by(team_display_name == team, season == year))
  
pre_deadline_stats <- playoff_joined%>%
  left_join(deadline_df, join_by(season == season))%>%
  filter(game_date < deadline_date)%>%
  mutate(minutes = if_else(is.na(minutes), 0, minutes))%>%
  group_by(athlete_display_name, season)%>%
  summarize(pre_deadline_min = mean(minutes))

joined_df <- playoff_joined%>%
  left_join(pre_deadline_stats, by = join_by(athlete_display_name == athlete_display_name,
                                             season == season))%>%
  mutate(pre_deadline_min = if_else(is.na(pre_deadline_min), 0, pre_deadline_min))%>%
  mutate(is_minutes_outlier = if_else(minutes > 20 & pre_deadline_min < 5, T, F))

summed_date <- joined_df%>%
  group_by(game_date)%>%
  summarize(total_min_outliers = sum(is_minutes_outlier, na.rm = T))

summed_team <- joined_df%>%
  group_by(team_display_name, season)%>%
  summarize(total_min_outliers = sum(is_minutes_outlier, na.rm = T))



summed_date%>%
  ggplot(aes(x = game_date, y = total_min_outliers))+
  geom_line()+
  theme_minimal()

container <- vector("list", length = length(seasons))
for (i in seq_along(seasons)){
  container[[i]] <- espn_nba_standings(seasons[i])%>%
    mutate(year = seasons[i])%>%
    select(year, team, wins)
}
standings <- bind_rows(container)

teams <- espn_nba_teams()%>%
  mutate(logo2 = logos_href_9)%>%
  select(display_name, logo, logo2)
  

xy <- summed_team%>%
  left_join(playoffs_filtered, by = join_by(team_display_name == team,
                                            season == year))%>%
  left_join(standings, by = join_by(season == year,
            team_display_name == team))%>%
  arrange(season, wins)%>%
  filter(!is.na(playoff_team))%>%
  add_column(rank = rep(30:1, 6))%>%
  left_join(teams, by = join_by(team_display_name == display_name))
```


## Write to csv

```{r}
write_csv(summed_team, "minutes_team.csv")
write_csv(summed_date, "minutes_date.csv")
write_csv(final_dat, "win_pct_slope_2026.csv")
write_csv(dat_calendar, "calendar_alltime.csv")
write_csv(final_dat_3, "10_day_avg_win_pct.csv")
write_csv(dat_3, "playoff_vs_non.csv")
write_csv(xy, "minutes_logo.csv")
```




